A Hexagon Tile Map Algorithm for Displaying Spatial Data

### 2020-10-12 kobakian-cook.tex
New tas displays

\abstract{%
This algorithm creates a tessellated hexagon display to represent a set
of the spatial polygons. It allocates these hexagon in a manner that
preserves the spatial relationship of the geographic units. The
algorithm creates a display that showcases spatial distributions, by
emphasising the small geographical regions that are often difficult to
locate on geographic maps. Spatial distributions have been presented on
alternative representations of geography for many years. In modern
times, interactivity and animation have begun to play a larger role, as
alternative representations have been popularised by online news sites,
and atlas websites with a focus on public consumption. Applications are
increasingly widespread, especially in the areas of disease mapping, and
election results.
}

@@

In many countries, residents are increasingly gathering to live in urban
areas around major cities \citep{ACTUC}. The creation of approximately
equal population areas results in dramatically different square meterage
of the administrative geographic units, such as states or electorates.
The units are filled with colour to represent the value of the statistic
for each unit \citep{EI}. This can lead to unequal emphasis on the
statistical value of each of the geographic units. This allows
choropleth map displays to misrepresent the spatial distributions of
human related statistics due to area-size bias \citep{BCM}. Figure
\ref{fig:melanoma-geo} is a choropleth map that uses colour to display
the estimated Standardised Incidence Ratios (SIRs) of melanoma cancer
for males for each of the Statistical Areas at Level 2 (SA2) used by the
\citet{abs2011}. The Australian choropleth map display draws attention
to the expanse of light blue areas across the rural communities in all
states. The SA2s around Hobart stand out as more orange and red. To
create this choropleth map the SA2 polygons for 2011 from the Australian
Bureau of Statistics.

The solutions to this visualisation problem begin with the geography.
Alternative maps shift the focus from land area and shape, to the value
of the statistics in a group of areas \citep{ACCAC}. Different styles of
cartograms apply transformations to the geographic areas, to highlight
the values of the value of the statistic of interest. These displays
result in a distortion of the map space to represent differences in the
statistic across the areas \citep{ACCAC} as the statistic of interest is
used to determine the cartogram layout.

\begin{figure}[h]
\centering
\includegraphics[width=14cm]{figs/aus_melanoma_p.png}
\caption{\label{fig:melanoma-geo}A choropleth map of the Statistical Areas of Australia at Level 2. The colours communicate the value of the estimated SIR of Melanoma for males, they range from much lower than average (blue) to much higher than average (red)}
\end{figure}

@@

There are several established alternative visualisation methods. Tile
maps, Rectangular cartograms \citep{ORC} and Dorling cartograms
\citep{ACTUC} all use one simple shape to represent each geographic
unit. They all minimise the emphasis on the size or shape of the
geographic areas. These alternative map displays focus on the
relationship between neighbours, attempting to preserve connections, and
disregard the unique shapes of the administrative boundaries. Figure
\ref{fig:nsw_grid} shows a collection of alternative map displays, this
includes a) a contiguous cartogram, b) a non-contiguous cartogram, c) a
Dorling cartogram and d) a hexagon tile map. Each Statistical Area at
Level 2 (SA2) \citep{abs2011} was designed to represent a community.
These administrative areas are often used to aggregate census data and
used to understand demographics of the Australian population.

The familiar choropleth map display is often used to allow the user to
orient themselves. However, the emphasis on the land mass diminishes the
features of the distribution in densely populated communities due to the
small size on the display \citep{ACTUC}. The unique shapes of boundaries
can be helpful for orienting users but may not contribute to their
understanding of the spatial disease distribution as many of the
communities are not visible in a choropleth display \citep{TVSSS}. This
presents an opportunity for explicit map transformations to improve
communication of distributions \citep{CBATCC}. In figure
\ref{fig:nsw_grid}a) the choropleth map is presented with the
alternative maps to allow for comparisons to be made.

\begin{Schunk}

\includegraphics[width=0.9\linewidth]{kobakian-cook_files/figure-latex/tas_displays-1} \end{Schunk}

When communicating information that is relevant to the population, each
member of the population can be given equal representation by
transforming the map \citep{TVSSS}. The impact of a disease over
connected communities can be seen by allowing the boundaries to maintain
connection in the transformed display. The contiguous cartogram
displayed in figure \ref{fig:nsw_grid}b) draws attention to smaller
geographic units when they are rescaled according to the population
\citep{DMAHP}. These new shapes can now be coloured to represent a
second variable. This display can create twisted and unfamiliar shapes
from the geographic units as the algorithms must satisfy the topology
conditions, especially when there are communities located geographically
far from their neighbours \citep{TVSSS}. This shown in the northern SA2s
of Tasmania with an average (yellow) estimated SIR of Melanoma for
males.

The non-contiguous cartogram in figure \ref{fig:nsw_grid}c) also uses
the population to rescale the geographic units. Unlike the contiguous
cartogram, the SA2 areas maintain their geographic shape, but they may
not retain the connection with their neighbours. The population of the
SA2 areas is used to scale the geographic units in the non-contiguous
cartogram \citep{NAC}. The amount of white space can be meaningful in
non-contiguous cartograms \citep{ECGC}, in this example the disparity
between large rural areas and small urban areas means that the city of
Hobart, shown in the inset map looks reasonable, but these units are not
visible in the context of the Tasmania state map. Depending on the
difference between the population and geographic land size, the amount
of white space can also prevent meaningful understanding of the
distribution \citep{TVSSS}.

The Dorling cartogram presents each geographic unit as a circle, the
size of the circle is scaled according to the population value of each
area \citep{ACTUC}. Figure \ref{fig:nsw_grid}d) shows the Tasmania SA2
areas as an individual circle located as close as possible to the
geographic centroid location. This map draws attention to the collection
of coastal cities in Tasmania that were not apparent in figure
\ref{fig:nsw_grid} a), b) or c).

@@

There are three key steps performed to create a tessellated hexagon tile
map. First derive the set of centroids from the polygons provided, then
create the grid of hexagons locations. These two steps are defined in
the blue left column of the flow chart in Figure
\ref{fig:sugarbag_flow}. Each centroid can then be allocated to an
available hexagon location. The steps for the allocation process are
detailed in the right column of Figure \ref{fig:sugarbag_flow}. To make
tessellated plots, the point locations are converted to hexagon shapes,
with the hexagon size used in the grid.

@@

\begin{figure}[h]
\centering
\includegraphics[width=14cm]{figs/1centroids.png}
\caption{\label{fig:centroids_plot}The geographic shapes of the Statistical Areas of Tasmania at Level 2. The points show the locations of the centroids of the SA2 areas.}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=16cm]{figs/2grid.png}
\caption{\label{fig:grid2}Grid points to create a hexagon tile map.}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=10cm]{figs/3grid.png}
\caption{\label{fig:filter-grid}The remaining hexagon locations from all the possible points in the initial grid after the buffer is applied. The blue dots show the grid points left to choose from after the buffer step.}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=16cm]{figs/4grid.png}
\caption{\label{fig:buffers}The remaining available hexagon locations after filtering for the grid points within a square distance, then circular around the centroid.}
\end{figure}

The \texttt{width} parameter is used to take a slice of the remaining
points. The width is the amount of degrees used on either side of the
angle from the focal point to centroid location. This uses the angle
from the closest capital city, to the current centroid as seen in Figure
\ref{fig:angles} . This allows the spatial relationship to be preserved,
even when it is allocated to a hexagon that is further from the focal
point then the original centroid location.

\begin{figure}[h]
\centering
\includegraphics[width=16cm]{figs/5allocate.png}
\caption{\label{fig:angles}The remaining available hexagon locations after filtering for grid points within the angle from the focal point to the centroid.}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=16cm]{figs/6allocate.png}
\caption{\label{fig:buffs}A complete hexagon tile map of Tasmania.}
\end{figure}

A complete hexagon tile map of Tasmania is created by applying the
algorithm steps to each centroid. The hexagon tile map visualisation is
used below to visualise the Australian Cancer Atlas data \citep{TACA}.
Two views of the same data are produced by filling according to the
thyroid Cancer Standardised Incidence Rates (SIRs) downloaded from the
Australian Cancer Atlas site. This small example in Figure \ref{fig:sir}
shows the group of blue areas in the Hobart CBD more prominently in the
hexagon tile map (b). The small red areas visible in the choropleth map
(a) along the north coast are much larger in the hexagon tile maps. The
hexagon tile map shows less yellow, this no longer overwhelms the map
space with the information regarding the rural areas.

\begin{figure}[h]
\centering
\includegraphics[width=14cm]{figs/7SIR.png}
\caption{\label{fig:sir}The Australian Cancer Atlas data displayed on a choropleth and hexagon tile map. The colour of each Statistical Area of Australian at Level 2 coloured by SIR. A choropleth map (a) of SIR is paired with a hexagon tile map (b) to contrast the colours that are made obvious when every SA2 is equally represented.}
\end{figure}

\hypertarget{neighbour-relationships}{%
\paragraph{Neighbour relationships}\label{neighbour-relationships}}

It is possible to consider the neighbouring areas for each SA2, for
stronger preservation of the spatial distribution.

An additional step can be included to allow the neighbours that have
already been allocated to influence the placement of the current
centroid. This requires specifying the \texttt{sf} object as the
argument for the \texttt{use\_neighbours} parameter. This calculates
neighbours using intersections of their polygons. This occurs for all
areas before any allocations begin.

During the allocation of each centroid, the list of neighbours is
consulted. If any neighbour was already allocated, the hexagons
surrounding the neighbours on the grid are prioritised. For multiple
neighbours, the neighbouring hexagon grid points are aggregated and
considered in order of distance from the original centroid.

@@

This algorithm has been applied to the complete set of all Australian
Statistical Areas at Level 2. This display highlights the density of
Australian capital cities, as it draws attention to the many communities
in Sydney, Melbourne and Hobart. This display also highlights the
disparity in the burden of Melanoma cancer for males in the communities
of these three cities. There are several collections of red hexagons in
Hobart that draw attention and they represent the communities with much
higher rates of diagnosis than the Australian average. The communities
south of Hobart show a gradient of colour as the values progress from
higher than average, to much lower than average for the communities
closer to the Sydney CBD. This pattern progresses into the communities
with lower than average rates in Melbourne and Tasmania.

@@

The Dorling cartogram was the most influential in the design of the
hexagon tile map. The use of a hexagon provides a simple shape that
shifts the focus of the visualisation from the shape of SA2s to the
distribution effects of a disease on communities. not impeded by the
irregular and unusual shapes created by a contiguous cartogram and the
immense white space created on the non-contiguous cartogram. Hexagons
can emphasise the relationship between neighbouring communities by
allowing their sides to touch. The tessellation crated by adjoining
hexagons for densely populated areas is helpful as it shows that most
communities are not islands onto themselves, but are intrinsically
connected to those around them. As seen in the Dorling cartogram in
figure \ref{fig:nsw_grid}c), the tessellation that allows boundary
connections was not imposed for rural communities, where centroid of the
geographic unit was located far from others, a feature that allows
distant but dense populations of the coastal towns of Tasmania to be
recognised.

The choropleth map that uses the geographic shapes of areas can be seen
in figure \ref{fig:melanoma-geo}. Figure \ref{fig:melanoma-hex} displays
the Australian areas with equally sized hexagons. This display allows
the small and densely populated inner city areas to be easily contrasted
with their neighbouring areas.

>>>>

\abstract{%
The algorithm presented creates a display that uses tessellated hexagons
to represent a set of spatial polygons. It allocates these hexagon in a
manner that preserves the spatial relationship of the geographic units,
in light of their positions to points of interest. The display showcases
spatial distributions, by emphasising the small geographical regions
that are often difficult to locate on geographic maps. Spatial
distributions have been presented on alternative representations of
geography for many years. In modern times, interactivity and animation
have allowed alternative displays to play a larger role. Alternative
representations have been popularised by online news sites, and atlas
websites with a focus on public consumption. Applications are
increasingly widespread, especially in the areas of disease mapping, and
election results.
}

@@

In many countries, residents are increasingly gathering to live in urban
areas around major cities \citep{ACTUC}. The creation of approximately
equal population areas results in dramatically different square meterage
of the administrative geographic units, such as states or electorates.
The units are filled with colour to represent the value of the statistic
for each unit \citep{EI}. This can lead to unequal emphasis on the
statistical value of each of the geographic units. This allows
choropleth map displays to misrepresent the spatial distributions of
human related statistics due to area-size bias \citep{BCM}. Figure
\ref{fig:choro} is a choropleth map that uses colour to display the
estimated Standardised Incidence Ratios (SIRs) of thyroid cancer for
males for each of the Statistical Areas at Level 2 (SA2) used by the
\citet{abs2011}. The Australian choropleth map display draws attention
to the expanse of dark and light blue areas across the rural communities
in all states. The SA2s around Brisbane stand out as more orange and
red.

\begin{Schunk}
\begin{figure}
\includegraphics[width=0.95\linewidth]{kobakian-cook_files/figure-latex/choro-1} \caption[A choropleth map of the Statistical Areas of Australia at Level 2]{A choropleth map of the Statistical Areas of Australia at Level 2. The colours communicate the value of the estimated SIR of Thyroid cancer for females, ranging from much lower than average (blue) to much higher than average (red)}\label{fig:choro}
\end{figure}
\end{Schunk}

The solutions to this visualisation problem begin with the geography.
Alternative maps shift the focus from land area and shape, to the value
of the statistics in a collection of areas \citep{ACCAC}. Each style of
cartogram applies a different transformation to the geographic areas, to
highlight the values of the statistic of interest. These displays result
in a distortion of the map space to represent features of the
distribution across the areas \citep{ACCAC} as the statistic of interest
is used to determine the cartogram layout.

@@

There are several established alternative visualisation methods. Tile
maps, Rectangular cartograms \citep{ORC} and Dorling cartograms
\citep{ACTUC} all use one simple shape to represent each geographic
unit. They all minimise the emphasis on the size or shape of the
geographic areas. These alternative map displays focus on the
relationship between neighbours, attempting to preserve connections, and
disregard the unique shapes of the administrative boundaries. Figure
\ref{fig:tas_displays} shows a collection of alternative map displays,
this includes a) a contiguous cartogram, b) a non-contiguous cartogram,
and c) a Dorling cartogram.

The familiar choropleth map display is often used to allow the user to
orient themselves. However, the emphasis on the land mass diminishes the
features of the distribution in densely populated communities due to the
small size on the display \citep{ACTUC}. The unique shapes of boundaries
can be helpful for orienting users but may not contribute to their
understanding of the spatial disease distribution as many of the
communities are not visible in a choropleth display \citep{TVSSS}. Each
Statistical Area at Level 2 (SA2) \citep{abs2011} was designed to
represent a community. These administrative areas are often used to
aggregate census data and used to understand demographics within
communities of the Australian population. This presents an opportunity
for explicit map transformations to improve communication of
distributions \citep{CBATCC}. In figure \ref{fig:tas_displays} the
choropleth map can be seen underneath each alternative map display to
allow for comparisons to be made.

\begin{Schunk}
\begin{figure}
\includegraphics[width=0.95\linewidth]{kobakian-cook_files/figure-latex/tas_displays-1} \caption[The three displays show alternative maps of the Statistical Areas at Level 2 in the Australian state of Tasmania]{The three displays show alternative maps of the Statistical Areas at Level 2 in the Australian state of Tasmania. These alternatives show the (a) contiguous cartogram, (b) non-contiguous cartogram and (c) Dorling cartogram of Tasmania.}\label{fig:tas_displays}
\end{figure}
\end{Schunk}

When communicating information that is relevant to the population, each
member of the population can be given equal representation by
transforming the map \citep{TVSSS}. The impact of a disease over
connected communities can be seen by allowing the boundaries to maintain
connection in the transformed display. The contiguous cartogram
displayed in figure \ref{fig:tas_displays}b) draws attention to smaller
geographic units when they are rescaled according to the population
\citep{DMAHP}. These new shapes can now be coloured to represent a
second variable. This display can create twisted and unfamiliar shapes
from the geographic units as the algorithms must satisfy the topology
conditions, especially when there are communities located geographically
far from their neighbours \citep{TVSSS}. This shown in the northern SA2s
of Tasmania with an average (yellow) estimated SIR of thyroid for males.

The non-contiguous cartogram in figure \ref{fig:tas_displays}c) also
uses the population to rescale the geographic units. Unlike the
contiguous cartogram, the SA2 areas maintain their geographic shape, but
they may not retain the connection with their neighbours. The population
of the SA2 areas is used to scale the geographic units in the
non-contiguous cartogram \citep{NAC}. The amount of white space can be
meaningful in non-contiguous cartograms \citep{ECGC}, in this example
the disparity between large rural areas and small urban areas means that
the city of Hobart, shown in the inset map looks reasonable, but these
units are not visible in the context of the Tasmania state map.
Depending on the difference between the population and geographic land
size, the amount of white space can also prevent meaningful
understanding of the distribution \citep{TVSSS}.

The Dorling cartogram presents each geographic unit as a circle, the
size of the circle is scaled according to the population value of each
area \citep{ACTUC}. Figure \ref{fig:tas_displays}c) shows the Tasmania
SA2 areas as an individual circle located as close as possible to the
geographic centroid location. This map draws attention to the collection
of coastal cities in Tasmania that were not apparent in figure
\ref{fig:tas_displays} a), or b).

@@

The result of the algorithm is a display that emphasises the densely
populated geographic regions surrounding the capital cities in
Australia. Each capital city forms it's own layer of connected hexagons.
Figure \ref{fig:hexmap} allows the distributions in the capital cities
to be viewed at a deeper scale. Sydney is now notable for the
considerable amount of red areas, not just for the surrounding orange
areas around the New South Wales capital visible in the choropleth
display in Figure \ref{fig:choro}. Brisbane and Perth also draw
attention for the many orange areas. Interestingly there is no clear
divide between rural and urban SA2 areas as Melbourne, Darwin, Adelaide
and Hobart have low rates of thyroid cancer diagnosis.

\begin{Schunk}
\begin{figure}
\includegraphics[width=0.95\linewidth]{kobakian-cook_files/figure-latex/hexmap-1} \caption[A hexagon tile map of the Statistical Areas of Australia at Level 2]{A hexagon tile map of the Statistical Areas of Australia at Level 2. The colours communicate the value of the estimated SIR of thyroid for males, they range from much lower than average (blue) to much higher than average (red).}\label{fig:hexmap}
\end{figure}
\end{Schunk}

There are three key steps performed to create the tessellated hexagon
tile map shown in Figure \ref{fig:hexmap}. First, derive the set of
centroids from the polygons provided, then create the grid of hexagons
locations. These two steps are defined in the blue left column of the
flow chart in Figure \ref{fig:sugarbag_flow}. Each centroid can then be
allocated to an available hexagon location. The steps for the allocation
process are detailed in the right column of Figure
\ref{fig:sugarbag_flow}. To make tessellated plots, the point locations
are converted to hexagon shapes, with the hexagon size used in the grid.

@@

\begin{Schunk}
\begin{figure}
\includegraphics[width=0.95\linewidth]{kobakian-cook_files/figure-latex/centroids_plot-1} \caption[The Australian Cancer Atlas data has determined the colour of each Statistical Area of Australian at Level 2]{The Australian Cancer Atlas data has determined the colour of each Statistical Area of Australian at Level 2. A choropleth map (a) of Standardised Incidence Rates (SIRs) is paired with a hexagon tile map (b) to contrast the colours that are made obvious when every SA2 is equally represented.}\label{fig:centroids_plot}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}
\includegraphics[width=0.95\linewidth]{kobakian-cook_files/figure-latex/grid2-1} \caption[Complete set of grid points to create a tile map]{Complete set of grid points to create a tile map.}\label{fig:grid2}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}
\includegraphics[width=0.95\linewidth]{kobakian-cook_files/figure-latex/grid3-1} \caption[The remaining hexagon locations from all the possible points in the initial grid after the buffer is applied]{The remaining hexagon locations from all the possible points in the initial grid after the buffer is applied. The blue dots show the grid points left to choose from after the buffer step.}\label{fig:grid3}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}
\includegraphics[width=1\linewidth]{kobakian-cook_files/figure-latex/buffers-1} \caption[Filter for grid points within a square, then circular, distance for those closest to the centroid]{Filter for grid points within a square, then circular, distance for those closest to the centroid.}\label{fig:buffers}
\end{figure}
\end{Schunk}

The \texttt{width} parameter is used to take a slice of the remaining
points. The width is the amount of degrees used on either side of the
angle from the focal point to centroid location. This uses the angle
from the closest capital city, to the current centroid as seen in Figure
\ref{fig:buffers2} . This allows the spatial relationship to be
preserved, even when it is allocated to a hexagon that is further from
the focal point then the original centroid location.

\begin{Schunk}
\begin{figure}
\includegraphics[width=1\linewidth]{kobakian-cook_files/figure-latex/buffers2-1} \caption[Filter for grid points within a square, then circular, distance for those closest to the centroid]{Filter for grid points within a square, then circular, distance for those closest to the centroid.}\label{fig:buffers2}
\end{figure}
\end{Schunk}

@@

A complete hexagon tile map of Tasmania is created by applying the
algorithm steps to each centroid. The hexagon tile map is used to
visualise the Australian Cancer Atlas data \citep{TACA}. Two views of
the same data are produced by filling according to the thyroid Cancer
Standardised Incidence Rates (SIRs) downloaded from the Australian
Cancer Atlas site. This small example in Figure \ref{fig:centroids_plot}
shows the group of blue areas in the Hobart CBD more prominently in the
hexagon tile map (b). The small red areas visible in the choropleth map
(a) along the north coast are much larger in the hexagon tile maps. The
hexagon tile map shows less yellow, this no longer overwhelms the map
space with the information regarding the rural areas.

@@

This algorithm has been applied to the complete set of all Australian
Statistical Areas at Level 2. This display highlights the density of
Australian capital cities, as it draws attention to the many communities
in Sydney, Melbourne and Hobart. This display also highlights the
disparity in the burden of thyroid cancer for males in the communities
of these three cities. There are several collections of red hexagons in
Hobart that draw attention and they represent the communities with much
higher rates of diagnosis than the Australian average. The communities
south of Hobart show a gradient of colour as the values progress from
higher than average, to much lower than average for the communities
closer to the Sydney CBD. This pattern progresses into the communities
with lower than average rates in Melbourne and Tasmania.

@@

The Dorling cartogram was the most influential in the design of the
hexagon tile map. The use of a hexagon provides a simple shape that
shifts the focus of the visualisation from the shape of SA2s to the
distribution effects of a disease on communities. not impeded by the
irregular and unusual shapes created by a contiguous cartogram and the
immense white space created on the non-contiguous cartogram. Hexagons
can emphasise the relationship between neighbouring communities by
allowing their sides to touch. The tessellation crated by adjoining
hexagons for densely populated areas is helpful as it shows that most
communities are not islands onto themselves, but are intrinsically
connected to those around them. As seen in the Dorling cartogram in
figure \ref{fig:tas_displays}c), the tessellation that allows boundary
connections was not imposed for rural communities, where centroid of the
geographic unit was located far from others, a feature that allows
distant but dense populations of the coastal towns of Tasmania to be
recognised.

The choropleth map that uses the geographic shapes of areas can be seen
in figure \ref{fig:choro}. Figure \ref{fig:hexmap} displays the
Australian areas with equally sized hexagons. This display allows the
small and densely populated inner city areas to be easily contrasted
with their neighbouring areas.
